Projecting genetic associations through gene expression patterns highlights disease etiology and drug mechanisms

### 2023-08-03 output/latex/manuscript.tex
add biorender note in figure 1
replace "Supplementary File X" by "Supplementary Data X"

Our first experiment attempted to answer whether genes in a disease-relevant LV could represent potential therapeutic targets.
For this, the first step was to obtain a set of genes strongly associated with a phenotype of interest.
Therefore, we performed a fluorescence-based CRISPR-Cas9 in the HepG2 cell line and identified 462 genes associated with lipid regulation (\protect\hyperlink{sec:methods:crispr}{Methods}).
From these, we selected two high-confidence gene sets that either caused a decrease or increase of lipids:
a lipids-decreasing gene-set with eight genes: \emph{BLCAP}, \emph{FBXW7}, \emph{INSIG2}, \emph{PCYT2}, \emph{PTEN}, \emph{SOX9}, \emph{TCF7L2}, \emph{UBE2J2};
and a lipids-increasing gene-set with six genes: \emph{ACACA}, \emph{DGAT2}, \emph{HILPDA}, \emph{MBTPS1}, \emph{SCAP}, \emph{SRPR} (Supplementary File 2).

@@

We found that phenotypes were grouped into five clear branches, defined by their first node at the top of the Figure \ref{fig:clustering:tree}:
0) a ``large'' branch that includes most of the traits subdivided only starting at \(k\)=16 (with asthma, subjective well-being traits, and nutrient intake clusters),
1) heel bone-densitometry measurements,
2) hematological assays on red blood cells,
3) physical measures, including spirometry and body impedance, and anthropometric traits with fat-free and fat mass measures in separate sub-branches, and
4) a ``complex'' branch including keratometry measurements, assays on white blood cells and platelets, skin and hair color traits, autoimmune disorders, and cardiovascular diseases (which also included other cardiovascular-related traits such as hand-grip strength\textsuperscript{\protect\hyperlink{ref-aBVh8zt1}{81}}, and environmental/behavioral factors such as physical activity and diet) (see Supplementary Files 3-7 for all clustering results).
Within these branches, results were relatively stable, with the same traits often clustered together across different resolutions.
Arrows between clusters show traits moving from one group to another, and this mainly happens between clusters within the ``complex'' branch (4) and between clusters from the ``large'' branch (0) to the ``complex'' branch.
This behavior is expected since complex diseases are usually associated with shared genetic and environmental factors and are thus hard to categorize into a single cluster.

@@

We used two large TWAS resources from different cohorts for discovery and replication, all obtained from European ancestries.
PhenomeXcan\textsuperscript{\protect\hyperlink{ref-lY5ln3dB}{42}}, our discovery cohort, provides results on 4,091 traits across different categories.
Supplementary File 1 has all the details about the included GWAS, sample size and disease/trait categories.
In PhenomeXcan, these publicly available GWAS summary statistics were used to compute
1) gene-based associations with the PrediXcan family of methods (described before), and
2) a posterior probability of colocalization between GWAS loci and \emph{cis}-eQTL with fastENLOC\textsuperscript{\protect\hyperlink{ref-lY5ln3dB}{42},\protect\hyperlink{ref-ndd3tW4g}{96}}.
We refer to the matrix of \(z\)-scores from S-PrediXcan (Equation (\ref{eq:spredixcan})) across \(q\) traits and \(m\) genes in tissue \(t\) as \(\mathbf{M}^{t} \in \mathbb{R}^{q \times m}\).
As explained later, matrices \(\mathbf{M}^{t}\) were used in our LV-based drug repurposing framework since they provide direction of effects.
The S-MultiXcan results (22,515 gene associations across 4,091 traits) were used in our LV-based regression framework and our cluster analyses of traits.
For the cluster analyses, we used the \(p\)-values converted to \(z\)-scores: \(\mathbf{M}=\Phi^{-1}(1 - p/2)\), where \(\Phi^{-1}\) is the probit function.
Higher \(z\)-scores correspond to stronger associations.

@@

\begin{figure}
\hypertarget{fig:entire_process}{%
\centering
% \includegraphics[width=1\textwidth,height=\textheight]{images/entire_process/entire_process.pdf}
\caption{\textbf{Schematic of the PhenoPLIER framework.}
\textbf{a)} High-level schematic of PhenoPLIER (a gene module-based method) in the context of TWAS (single-gene) and GWAS (single-variant).
In GWAS, we identify variants associated with traits.
In TWAS, first, we identify variants that are associated with gene expression levels (eQTLs); then, prediction models based on eQTLs are used to impute gene expression, which is used to compute gene-trait associations.
Resources such as LINCS L1000 provide information about how a drug perturbs gene expression; at the bottom-right corner, we show how a drug downregulates two genes (A and C).
In PhenoPLIER, these data types are integrated using groups of genes co-expressed across one or more conditions (such as cell types) that we call gene modules or latent variables/LVs.
\textbf{b)} The integration process in PhenoPLIER uses low-dimensional representations (matrices \(\mathbf{Z}\) and \(\mathbf{B}\)) learned from large gene expression datasets (top).
We used gene-drug information \(\mathbf{L}\) from LINCS L1000 and gene-trait associations \(\mathbf{M}\) from TWAS: PhenomeXcan was used as the discovery cohort, and eMERGE as replication (middle).
PhenoPLIER provides three computational components (bottom):
1) an LV-based regression model that associates an LV \(j\) (\(\mathbf{Z}_j\)) with a trait \(i\) (\(\mathbf{M}_i\)),
2) a clustering framework that learns groups of traits from TWAS associations projected into the LV space (\(\hat{\mathbf{M}}\)),
and 3) an LV-based drug repurposing approach that uses the projection of TWAS (\(\hat{\mathbf{M}}\)) and LINCS L1000 (\(\hat{\mathbf{L}}\)) into the LV space.
\textbf{c)} Genes that are part of LV603, termed as a neutrophil signature\textsuperscript{\protect\hyperlink{ref-14rnBunuZ}{44}}, were expressed in relevant cell types (top), with 53 independent samples expressed in Neutrophils, 59 in Granulocytes, and 20 in Whole blood, 56 in PBMC, 8 in mDCs, 29 in Monocytes, and 5 in Epithelial cells (the boxplot shows the 25th, 50th and 75th percentiles while the whiskers extend to the minimum/maximum values).
LV603 was associated in PhenoPLIER with neutrophil counts and other white blood cells (bottom, showing the top 10 traits for LV603 after projecting gene-trait associations in PhenomeXcan).
eQTLs: expression quantitative trait loci;
MVN: multivariate normal distribution;
PBMC: peripheral blood mononuclear cells;
mDCs: myeloid dendritic cells.}\label{fig:entire_process}
}
\end{figure}

@@

\begin{figure}
\hypertarget{fig:clustering:tree}{%
\centering
% \includesvg[width=1\textwidth,height=\textheight]{images/clustering/clustering_tree.svg}
\caption{\textbf{Clustering tree using multiple resolutions for clusters of traits.}
Each row represents a partition/grouping of the traits, and each circle is a cluster from that partition.
The number of clusters goes from 5 to 29.
Arrows indicate how traits in one cluster move across clusters from different partitions.
Most of the clusters are preserved across different resolutions, showing highly stable solutions even with independent runs of the clustering algorithm.
RDW: red cell (erythrocyte) distribution width;
BMI: body mass index;
WC: waist circumference;
HC: hip circumference;
RA: rheumatoid arthritis;
SLE: systemic lupus erythematosus;
HTN: Hypertension;
IBD: inflammatory bowel disease;
SCZ: Schizophrenia;
CAD: Coronary artery disease;
AD: Alzheimer's disease;
The full lists of traits in each cluster in the last five partitions of the tree (from \(k=16\) to \(k=29\)) are in Supplementary Files 3-7.}\label{fig:clustering:tree}
}
\end{figure}

>>>>

Our first experiment attempted to answer whether genes in a disease-relevant LV could represent potential therapeutic targets.
For this, the first step was to obtain a set of genes strongly associated with a phenotype of interest.
Therefore, we performed a fluorescence-based CRISPR-Cas9 in the HepG2 cell line and identified 462 genes associated with lipid regulation (\protect\hyperlink{sec:methods:crispr}{Methods}).
From these, we selected two high-confidence gene sets that either caused a decrease or increase of lipids:
a lipids-decreasing gene-set with eight genes: \emph{BLCAP}, \emph{FBXW7}, \emph{INSIG2}, \emph{PCYT2}, \emph{PTEN}, \emph{SOX9}, \emph{TCF7L2}, \emph{UBE2J2};
and a lipids-increasing gene-set with six genes: \emph{ACACA}, \emph{DGAT2}, \emph{HILPDA}, \emph{MBTPS1}, \emph{SCAP}, \emph{SRPR} (Supplementary Data 2).

@@

We found that phenotypes were grouped into five clear branches, defined by their first node at the top of the Figure \ref{fig:clustering:tree}:
0) a ``large'' branch that includes most of the traits subdivided only starting at \(k\)=16 (with asthma, subjective well-being traits, and nutrient intake clusters),
1) heel bone-densitometry measurements,
2) hematological assays on red blood cells,
3) physical measures, including spirometry and body impedance, and anthropometric traits with fat-free and fat mass measures in separate sub-branches, and
4) a ``complex'' branch including keratometry measurements, assays on white blood cells and platelets, skin and hair color traits, autoimmune disorders, and cardiovascular diseases (which also included other cardiovascular-related traits such as hand-grip strength\textsuperscript{\protect\hyperlink{ref-aBVh8zt1}{81}}, and environmental/behavioral factors such as physical activity and diet) (see Supplementary Data 3-7 for all clustering results).
Within these branches, results were relatively stable, with the same traits often clustered together across different resolutions.
Arrows between clusters show traits moving from one group to another, and this mainly happens between clusters within the ``complex'' branch (4) and between clusters from the ``large'' branch (0) to the ``complex'' branch.
This behavior is expected since complex diseases are usually associated with shared genetic and environmental factors and are thus hard to categorize into a single cluster.

@@

We used two large TWAS resources from different cohorts for discovery and replication, all obtained from European ancestries.
PhenomeXcan\textsuperscript{\protect\hyperlink{ref-lY5ln3dB}{42}}, our discovery cohort, provides results on 4,091 traits across different categories.
Supplementary Data 1 has all the details about the included GWAS, sample size and disease/trait categories.
In PhenomeXcan, these publicly available GWAS summary statistics were used to compute
1) gene-based associations with the PrediXcan family of methods (described before), and
2) a posterior probability of colocalization between GWAS loci and \emph{cis}-eQTL with fastENLOC\textsuperscript{\protect\hyperlink{ref-lY5ln3dB}{42},\protect\hyperlink{ref-ndd3tW4g}{96}}.
We refer to the matrix of \(z\)-scores from S-PrediXcan (Equation (\ref{eq:spredixcan})) across \(q\) traits and \(m\) genes in tissue \(t\) as \(\mathbf{M}^{t} \in \mathbb{R}^{q \times m}\).
As explained later, matrices \(\mathbf{M}^{t}\) were used in our LV-based drug repurposing framework since they provide direction of effects.
The S-MultiXcan results (22,515 gene associations across 4,091 traits) were used in our LV-based regression framework and our cluster analyses of traits.
For the cluster analyses, we used the \(p\)-values converted to \(z\)-scores: \(\mathbf{M}=\Phi^{-1}(1 - p/2)\), where \(\Phi^{-1}\) is the probit function.
Higher \(z\)-scores correspond to stronger associations.

@@

\begin{figure}
\hypertarget{fig:entire_process}{%
\centering
% \includegraphics[width=1\textwidth,height=\textheight]{images/entire_process/entire_process.pdf}
\caption{\textbf{Schematic of the PhenoPLIER framework.}
\textbf{a)} High-level schematic of PhenoPLIER (a gene module-based method) in the context of TWAS (single-gene) and GWAS (single-variant).
In GWAS, we identify variants associated with traits.
In TWAS, first, we identify variants that are associated with gene expression levels (eQTLs); then, prediction models based on eQTLs are used to impute gene expression, which is used to compute gene-trait associations.
Resources such as LINCS L1000 provide information about how a drug perturbs gene expression; at the bottom-right corner, we show how a drug downregulates two genes (A and C).
In PhenoPLIER, these data types are integrated using groups of genes co-expressed across one or more conditions (such as cell types) that we call gene modules or latent variables/LVs. Created with BioRender.com.
\textbf{b)} The integration process in PhenoPLIER uses low-dimensional representations (matrices \(\mathbf{Z}\) and \(\mathbf{B}\)) learned from large gene expression datasets (top).
We used gene-drug information \(\mathbf{L}\) from LINCS L1000 and gene-trait associations \(\mathbf{M}\) from TWAS: PhenomeXcan was used as the discovery cohort, and eMERGE as replication (middle).
PhenoPLIER provides three computational components (bottom):
1) an LV-based regression model that associates an LV \(j\) (\(\mathbf{Z}_j\)) with a trait \(i\) (\(\mathbf{M}_i\)),
2) a clustering framework that learns groups of traits from TWAS associations projected into the LV space (\(\hat{\mathbf{M}}\)),
and 3) an LV-based drug repurposing approach that uses the projection of TWAS (\(\hat{\mathbf{M}}\)) and LINCS L1000 (\(\hat{\mathbf{L}}\)) into the LV space.
\textbf{c)} Genes that are part of LV603, termed as a neutrophil signature\textsuperscript{\protect\hyperlink{ref-14rnBunuZ}{44}}, were expressed in relevant cell types (top), with 53 independent samples expressed in Neutrophils, 59 in Granulocytes, and 20 in Whole blood, 56 in PBMC, 8 in mDCs, 29 in Monocytes, and 5 in Epithelial cells (the boxplot shows the 25th, 50th and 75th percentiles while the whiskers extend to the minimum/maximum values).
LV603 was associated in PhenoPLIER with neutrophil counts and other white blood cells (bottom, showing the top 10 traits for LV603 after projecting gene-trait associations in PhenomeXcan).
eQTLs: expression quantitative trait loci;
MVN: multivariate normal distribution;
PBMC: peripheral blood mononuclear cells;
mDCs: myeloid dendritic cells.}\label{fig:entire_process}
}
\end{figure}

@@

\begin{figure}
\hypertarget{fig:clustering:tree}{%
\centering
% \includesvg[width=1\textwidth,height=\textheight]{images/clustering/clustering_tree.svg}
\caption{\textbf{Clustering tree using multiple resolutions for clusters of traits.}
Each row represents a partition/grouping of the traits, and each circle is a cluster from that partition.
The number of clusters goes from 5 to 29.
Arrows indicate how traits in one cluster move across clusters from different partitions.
Most of the clusters are preserved across different resolutions, showing highly stable solutions even with independent runs of the clustering algorithm.
RDW: red cell (erythrocyte) distribution width;
BMI: body mass index;
WC: waist circumference;
HC: hip circumference;
RA: rheumatoid arthritis;
SLE: systemic lupus erythematosus;
HTN: Hypertension;
IBD: inflammatory bowel disease;
SCZ: Schizophrenia;
CAD: Coronary artery disease;
AD: Alzheimer's disease;
The full lists of traits in each cluster in the last five partitions of the tree (from \(k=16\) to \(k=29\)) are in Supplementary Data 3-7.}\label{fig:clustering:tree}
}
\end{figure}
